---
title: "Starting with with the dplyr, ggplot2 packages, and the pipe operator |>"
author: "Han Olff"
date: "2024-08-29"
format: html
editor: 
  markdown: 
    wrap: 72
editor_options:
  chunk_output_type: console
---

# Working with the dplyr and ggplot2 packages

Load the required libraries :

```{r message = FALSE, warning = FALSE}
# restore your library (keep at the start of every script)
renv::restore()

# clear the R environment
rm(list = ls())

# load the required packages
library(tidyverse)  # this loads a group of related packages, including readr, dplyr and ggplot2
```

We will work with an online database of the Schiermonnikoog transect
study, that you will collect additional data for next week in the field.
In this example, we work with measurements of cockles (a bivalve
mollusc) on their width and length. From the database Epibenthos, we
will work with the table
[FactCockles](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1538766002#gid=1538766002).
See the documentation of the different variables in the table
[MetVariables](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1290622213#gid=1290622213)

### Read the datafile from the Google sheets database

To read the data in R, you first need to know the published csv link of
the FactCockles table. In the database, you can find this link in the
table MetTables. It is produced in Google Sheets throught the menu
File/Share/Publish to web and then selecting the table and output as
csv.

Read the FactCockleSize table with read_csv, which reads it as tibble
(formatted dataframe)

```{r}
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv")
print(FactCockleSize)
names(FactCockleSize)
nrow(FactCockleSize)

```

Plot with ggplot the relation between cockle thickness (thickness_mm, as
x) and length (length_mm), showing each year with a different color, and
add a regression line through all the years

```{r}
ggplot2::ggplot() # show what you are using
# own try
ggplot(data=FactCockleSize |> dplyr::filter(thickness_mm <300), 
       mapping = aes(x = thickness_mm,
                     y = length_mm, color = year)) +
  geom_point() +
geom_smooth(method = 'lm', col = 'purple' )   ##Regression line



# in class 
ggplot2::ggplot(data=FactCockleSize,  #what data table to use
                mapping=aes(x=length_mm,y=thickness_mm))  #what is on the axes

# still need to choose type of graph 
ggplot2::ggplot(data=FactCockleSize,  
                mapping=aes(x=length_mm,y=thickness_mm)) + 
  geom_point()

# A big outlier is seen. Likely a data entry mistake, remove it (not likely 1,5m cockle with 25mm length)
FactCockleSize |> dplyr::filter(thickness_mm>500)  #know obs_ID
# want to make a new dataset (but want to do it as little as possible) read data again without the mistaken observation

FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv") |> 
  dplyr::filter(CockleObs_ID!=1531, CockleObs_ID!=469) |>
  dplyr::mutate(year=factor(year))   #make year a factor
print(FactCockleSize)

# plot data again without outlier
ggplot2::ggplot(data=FactCockleSize,  
                mapping=aes(x=length_mm,y=thickness_mm)) + 
  geom_point()

# find other outlier
FactCockleSize |> dplyr::filter(thickness_mm>10 & length_mm<5)
# also remove observation 469 above
```

Further explore the plot with a regression line through all the data

```{r}
# own try
ggplot(data=FactCockleSize |> dplyr::filter(thickness_mm <300), 
       mapping = aes(x = thickness_mm,
                     y = length_mm, color = TransectPoint_ID)) +
  geom_point() +
geom_smooth(method = 'lm', col = 'red' )



# in class, with a regression line
ggplot2::ggplot(data=FactCockleSize,  
                mapping=aes(x=length_mm,y=thickness_mm)) + 
  geom_point() +
  geom_smooth(method="lm")

#calculate a linear regression
model_lm<-lm(thickness_mm~length_mm, data=FactCockleSize) 
summary(model_lm)

# equation: thickness_mm=-0.353+0.699*length_mm
# a cockle of lenth 20 mm has an predicted width of:
-0.353+0.699*20 #13.627
# this regression is significant, R2=0.9, P<0.001 (***)
```

Make same plot but showing a separate regression line per year

```{r}
# color the points by year, but plot one regression line
ggplot(data=FactCockleSize |> dplyr::filter(thickness_mm <300), 
       mapping = aes(x = thickness_mm,
                     y = length_mm, color = as.factor(year))) +
  geom_point() +
geom_smooth( method = "lm", se = F)+
  theme_minimal()


# in class
ggplot2::ggplot(data=FactCockleSize,  
                mapping=aes(x=length_mm,y=thickness_mm)) + 
  geom_point(mapping=aes(col=year)) +
  geom_smooth(method="lm")
```

Make a panel plot where with each year is shown as a separate graph

```{r}
ggplot(data=FactCockleSize |> dplyr::filter(thickness_mm <300), 
       mapping = aes(x = thickness_mm,
                     y = length_mm, color = as.factor(year))) +
  geom_point() +
  facet_wrap(~ year) +
geom_smooth(method = 'lm', col = 'blue' )


# in class
drop_na(FactCockleSize) |>
  ggplot2::ggplot(mapping=aes(x=length_mm,y=thickness_mm)) + 
    geom_point(mapping=aes(col=year), size=2) +
    xlab("cockle length (mm)") + 
    ylab("cockle width (mm)") + 
    geom_smooth(method="lm") +
    facet_wrap(~year)

# correlation found, not much difference between years. Length is a fine indicator for biomass
```

We conclude from this that: \* there were two important outliers in the
dataset that were removed after visual inspection \* the regression
between length and width is abou the same for every year, we safely can
use only length as a proxy for the biomass of an individual cockle
